Approximation as Differentiation: Notes

### 2025-05-21 main/formal-structures.tex
Tweak.

\noindent The trivial 1-point lattice, which is both terminal and initial, is the unit for $\biprod$. Although
the presence of biproducts is incompatible with $\LatGal$ itself being Cartesian closed, the biproduct
structure will prove crucial for establishing a Cartesian closed setting for interpreting higher-order
programs for Galois slicing.

@@

\begin{definition}[Semi-additive category]
A category with finite products and coproducts is \emph{semi-additive} if the canonical morphisms (projections
and injections) give an isomorphism
\[\textstyle X \coprod Y \iso X \prod Y\] that is natural in both variables.
\end{definition}

@@

\subsubsection{Biproduct laws}
It is also easy to show that the following equations hold:

@@

Now consider the total category $\Fam(\LatGal)$, which is our basic setting for interpreting first-order
programs for Galois slicing. This has as objects $(X, \partial X)$, all pairs of a set $X$ and and for every
$x \in X$, a bounded lattice $\partial X(x)$, and as morphisms $(X, \partial X) \to (Y, \partial Y)$, all
pairs $(f, \partial f)$ of a function $f: X \to Y$ and for every $x \in X$, a Galois connection $\partial
f(x): \partial X(x) \to \partial Y(f(x))$, as required. The pullback (reindexing) along $f$ in the composition
$(g, \partial g) \comp (f, \partial f) = (g \comp f, \reindex{\partial g}{f} \comp \partial f)$ selects the
appropriate lattice of approximations and Galois connection at each point, using the baseline (unapproximated)
function $f$.

>>>>

\noindent The trivial 1-point lattice, which is both terminal and initial and we will write as $0$, is the
unit for $\biprod$. Such a structure $(\biprod, 0)$ is called a \emph{biproduct} (\secref{biproducts}).
Although $\LatGal$ is not itself Cartesian closed (in particular the set of Galois connections $X \to Y$ does
not in general form a bounded lattice), the biproduct structure will prove crucial for establishing a
Cartesian closed setting for interpreting higher-order programs.

@@

\begin{definition}[Semi-additive category]
\label{def:biproducts:semi-additive}
A category with finite products and coproducts is \emph{semi-additive} if the canonical morphisms (projections
and injections) give an isomorphism
\[\textstyle X \coprod Y \iso X \prod Y\] that is natural in both variables.
\end{definition}

@@

\subsubsection{Biproduct laws}
It is also easy to show that in a semi-additive category, as defined in \defref{biproducts:semi-additive}, the
following equations hold:

@@

Now consider the total category $\Fam(\LatGal)$, which is our basic setting for interpreting first-order
programs for Galois slicing. This has as objects $(X, \partial X)$, all pairs of a set $X$ and and for every
$x \in X$, a bounded lattice $\partial X(x)$, and as morphisms $(X, \partial X) \to (Y, \partial Y)$, all
pairs $(f, \partial f)$ of a function $f: X \to Y$ and for every $x \in X$, a Galois connection $\partial
f(x): \partial X(x) \to \partial Y(f(x))$, as we sketched informally at the beginning of this section. The
pullback (reindexing) along $f$ in the composition $(g, \partial g) \comp (f, \partial f) = (g \comp f,
\reindex{\partial g}{f} \comp \partial f)$ selects the appropriate lattice of approximations and Galois
connection at each point, using the baseline (unapproximated) function $f$.
